The yaq Project: \\ A Standard for Scientific Instrumentation

### 2022-11-01 paper.tex
Variety of smaller fixes to spelling and grammar/word choice (largely) suggested by JCW and DDK

\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1} % consider using RevTex4.2

@@

Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
Increasing modular, commercial hardware availability has brought greater instrument complexity to small research groups.
The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional, monolithic orchestration software.
In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision even when highly modular hardware components enable extreme experimental flexibility.
Herein, we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of complex in
\yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
A large number of hardware peripherals are already supported with tooling available to create more.
The \yaq{} standard enables collaboration among multiple research groups, increasing code quality while lowering development effort.

@@

Instrumentation development is a key part of the scientific enterprise.
Novel instruments are typically constructed of many individual components that are both purchased and home-built.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
The challenge of integrating all of these interfaces is a frustrating piece of the modern instrument development process.
Weeks can be spent just integrating one new component into an existing project.
In small academic labs, these software interfaces are typically created by student researchers without software development experience.
Student researchers rarely focus on software reusability, and a lack of maitenence and documentation can make such software more difficult to use as time goes on.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

Some large user-facilities have addressed interface complexity via the adoption of unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
The unified standards define a network interface for any hardware component.
Orchestration software can target these unified standards for reading and writing hardware state.
Small background services are written to translate the myriad component interfaces into the standard EPICS IOCs and TANGO Devices.
These programs are performant, open source, and have huge libraries of exisiting hardware interface support, but require expert management to set up and provide descriptions via a separate server program.
In our experience EPICS and TANGO do not scale well to single-investigator lab environments.

@@

We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
We have built this standard to be self-describing, portable, and reusable wherever possible.
Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.

Here, we discuss the design of the \yaq{} standard in the context of four challenges facing instrument designers.
First, we discuss how particularly challenging hardware interfaces can become seemingly insurmountable barriers to software control.
Next, we discuss how inflexible orchestration software can limit experimental creativity.
Then, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
Finally, we discuss the heavy software maintenance burden that many instrument designers face.
In each case, we will highlight how the \yaq{} project is designed to alleviate that challenge.
Several case studies provide a view into the flexible ways that \yaq{} can be applied to perform different scientific experiments.
This paper provides an overview of the concepts and motivations behind \yaq{}.
Refer to the \yaq{} website for a formal specification of the \yaq{} standard.\cite{yaq_fyi}

@@

Figure \ref{fig:network} diagrams the \yaq{} architecture.
Here, we show three different computers connected via an ethernet network.
The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
This might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
That same computer is running four separate programs, one for each peripheral.
These small, targeted, programs are managed by the operating system and run in the background.
It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
The middle computer is connected to two additional peripherals, and runs daemons for each.
Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
The four client programs shown in Figure \ref{fig:network} can each communicate with all six hardware peripherals shown.
As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
This powerful architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
This client-server architecture offers similar network capabilities to EPICS and TANGO.
As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.

In \yaq{}, communication between daemons and clients is performed over TCP/IP using Apache Avro RPC \cite{AvroSpecification}.
Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote (client) process.
Practically, the \yaq{} interface looks like a collection of methods or functions, which Avro calls ``messages''.
Each message has defined input parameters and output return types.
A sensor might implement a message called ``\texttt{get\_measured}'' which takes no parameters and returns a dictionary mapping channel names to numeric or array data.
A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null} '' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
These self describing messages make up the lowest level functionality of the \yaq{} interface.
Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.

@@

The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system. % consider making this sentence (and the next paragraph starts) subsection headings 
\yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
The trait system was introduced in pursuit of our primary goal of easing the client development process.
Clients can trust that protocols that implement a given trait will behave in similar ways.
The standardized \yaq{} interface presents the same set of interactions for client-side scientific code, simplifying the experience of using hardware.

@@

Unique experiments will always need custom orchestration and user experience.
We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
Developing experimental software is an iterative process tied to the scientific goals of the instrument.
Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a} % citation does not compile
\yaq{} is architected to encourage better software development practices when creating such programs.
In a \yaq{} context, orchestration code and graphical control interfaces are implemented as clients.
These clients are automatically simpler because they only need to implement the \yaq{} standard and do not need to include the vast array of hardware-specific communication interfaces.
Beyond this, clients can use traits to interact with similar hardware identically.
A client written to perform a two dimensional fluorescence experiment using two Acton monochromators will also work with Horiba monochromators without any modification.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/client_spectrum},
	\caption{  \label{fig:foundation} \yaq{} scales from simple scripts to integrations with featureful frameworks.}
\end{figure}

A \yaq{} client can scale in complexity from a small, lightweight script all the way to a complicated graphical program, as shown in Figure \ref{fig:foundation}.
At the bottom, we represent the most lightweight interface to \yaq{}, Python scripting.
\texttt{yaqc}\cite{yaqc} is a Python client which is excellent for using in scripts or any other Python program.
The code shown creates three client objects which could be used directly through an interactive Python prompt or in a reusable script.
Each client object provides Python methods for each Avro message specified by the associated protocol.

@@

All three types of clients represented in Figure \ref{fig:foundation} have been shown addressing the same three hardware peripherals.
All types of clients can be used simultaneously to interact with the same instrument in different modes.
Client sophistication can be introduced naturally, as novel experiments are tested and refined.
Different clients can specialize for different requirements of a custom instrument
For example, \texttt{yaqc-qtpy} can provide a quick interface for setting and viewing hardware positions while Bluesky can focus on experimental data acquisition.

@@

The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
Remote client development on viritual hardware has allowed client design to proceed without delaying ongling experiments.
Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
Moving forward, the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figures/lines_histogram},
	\caption{  \label{fig:histogram} Histogram of the number of lines for each implemented daemon.
	Some daemons are implemented in ways that share code, resulting in apparent line counts less than 10.
	For example, the Thorlabs APT\cite{thorlabs_apt} motor implementation supports at least eight different daemons with each specifying only a handful of constants.
	These are extreme examples which are not representative of most hardware interfaces, so we omit them here. % it's not clear to me why these instances are not included (and scaled down???) It is a _feature_ not a bug that one can implement motor support with tiny amounts of code.
	}
\end{figure}

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.

While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in fewer than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{}.
In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
There are currently 70 daemons in the yaq project supporting at least 50 kinds of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}

@@

Software documentation is famously difficult and thankless work.
\yaq{} attempts to automate daemon documentation as much as possible.
Our website, https://yaq.fyi, automatically builds generated reference pages for all known protocols.
These pages are automatically updated when new versions are published.

@@

\section*{References}
% all yaq,yaqc,and yaqd citations have an unfilled "()" in them... 
\nocite{*}
\bibliography{references}

>>>>

\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}

@@

Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
Increasing hardware availability has brought greater instrument complexity to small research groups.
The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional monolithic orchestration software.
In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision.
Highly modular hardware components enable extreme experimental flexibility.
Herein we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of such instruments.
\yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
A large number of hardware peripherals are already supported with tooling available to create more.
The \yaq{} standard enables collaboration between multiple research groups, increasing code quality while lowering development effort.

@@

Instrumentation development is a key part of the scientific enterprise.
Novel instruments are typically constructed of many individual components, purchased and home-built.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
Weeks can be spent just integrating one new component into an existing project.
In small academic labs, these software interfaces are typically created by student researchers without software development experience.
Student researchers rarely focus on software reusability, and a lack of maintenance and documentation can make such software more difficult to use as time goes on.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

Some large user-facilities have addressed interface complexity via the adoption of unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
The unified standards define a network interface for any hardware component.
Orchestration software can target these unified standards for reading and writing hardware state.
Small background services are written to translate the myriad component interfaces into the standard EPICS IOCs and TANGO Devices.
These programs are performant, open source, and have huge libraries of existing hardware interface support, but require expert management to set up and provide descriptions via a separate server program.
In our experience EPICS and TANGO do not scale well to single-investigator lab environments.

@@

We have created a new network-based standard for scientific instrumentation, \yaq{}.
This standard borrows the most important ideas from established projects used by large user facilities while retaining simplicity appropriate for small research labs.
We have built this standard to be self-describing, portable, and reusable wherever possible.
Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.

Here, we discuss the design of the \yaq{} standard in the context of four challenges facing instrument designers.
First, we discuss how particularly challenging hardware interfaces can become seemingly insurmountable barriers to software control.
Next, we discuss how inflexible orchestration software can limit experimental creativity.
Next, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
Finally, we discuss the heavy software maintenance burden that many instrument designers face.
In each case, we will highlight how the \yaq{} project is designed to alleviate that challenge.
Several case studies provide a view into the flexible ways that \yaq{} can be applied to perform different scientific experiments.
This paper provides an overview of the concepts and motivations behind \yaq{}.
Refer to the \yaq{} website for a formal specification of the \yaq{} standard.\cite{yaq_fyi}

@@

Figure \ref{fig:network} diagrams the \yaq{} architecture.
Here, we show three different computers connected via an Ethernet network.
The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
This might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
That same computer is running four separate programs, one for each peripheral.
These small, targeted, programs are managed by the operating system and run in the background.
It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
The middle computer is connected to two additional peripherals, and runs daemons for each.
Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
The four client programs shown in Figure \ref{fig:network} can each communicate with all six hardware peripherals shown.
As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
This powerful architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
This client-server architecture offers similar network capabilities to EPICS and TANGO.
As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.

In \yaq{}, communication between daemons and clients is performed over TCP/IP using Apache Avro RPC \cite{AvroSpecification}.
Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote (client) process.
Practically, the \yaq{} interface looks like a collection of methods or functions, which Avro calls ``messages''.
Each message has defined input parameters and output return types.
A sensor might implement a message called ``\texttt{get\_measured}'' which takes no parameters and returns a dictionary mapping channel names to numeric or array data.
A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null}'' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
These self describing messages make up the lowest level functionality of the \yaq{} interface.
Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.

@@

The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.
\yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
The trait system was introduced in pursuit of our primary goal of easing the client development process.
Clients can trust that protocols that implement a given trait will behave in similar ways.
The standardized \yaq{} interface presents the same set of interactions for client-side scientific code, simplifying the experience of using hardware.

@@

Unique experiments will always need custom orchestration and user experience.
We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
Developing experimental software is an iterative process tied to the scientific goals of the instrument.
Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a}
\yaq{} is architected to encourage better software development practices when creating such programs.
In a \yaq{} context, orchestration code and graphical control interfaces are implemented as clients.
These clients are automatically simpler because they only need to implement the \yaq{} standard and do not need to include the vast array of hardware-specific communication interfaces.
Beyond this, clients can use traits to interact with similar hardware identically.
A client written to perform a two dimensional fluorescence experiment using two Acton monochromators will also work with Horiba monochromators without any modification.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/client_spectrum}
	\caption{  \label{fig:foundation} \yaq{} scales from simple scripts to integrations with featureful frameworks.}
\end{figure}

A \yaq{} client can scale in complexity from a small, lightweight script all the way to a sophisticated graphical program, as shown in Figure \ref{fig:foundation}.
At the bottom, we represent the most lightweight interface to \yaq{}, Python scripting.
\texttt{yaqc}\cite{yaqc} is a Python client which is excellent for using in scripts or any other Python program.
The code shown creates three client objects which could be used directly through an interactive Python prompt or in a reusable script.
Each client object provides Python methods for each Avro message specified by the associated protocol.

@@

All three types of clients represented in Figure \ref{fig:foundation} have been shown addressing the same three hardware peripherals.
All types of clients can be used simultaneously to interact with the same instrument in different modes.
Client sophistication can be introduced naturally, as novel experiments are tested and refined.
Different clients can specialize for different requirements of a custom instrument.
For example, \texttt{yaqc-qtpy} can provide a quick interface for setting and viewing hardware positions while Bluesky can focus on experimental data acquisition.

@@

The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
Remote client development on virtual hardware has allowed client design to proceed without delaying ongoing experiments.
Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
Moving forward, the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figures/lines_histogram},
	\caption{  \label{fig:histogram} Histogram of the number of lines for each implemented daemon.
	Some daemons are implemented in ways that share code, resulting in apparent line counts less than 10.
	For example, the Thorlabs APT\cite{thorlabs_apt} motor implementation supports at least eight different daemons with each specifying only a handful of constants.
	These are extreme examples which are not representative of most hardware interfaces, so we omit them here.
	}
\end{figure}

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in as less than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{}.
In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
There are currently 70 daemons in the yaq project supporting at least 50 kinds of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}

@@

Software documentation is famously difficult and thankless work.
\yaq{} attempts do automate daemon documentation as much as possible.
Our website, https://yaq.fyi, automatically builds generated reference pages for all known protocols.
These pages are automatically updated when new versions are published.

@@

\section*{References}

\nocite{*}
\bibliography{references}
